Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD

### 2022-09-15 paper/paper.qmd

---
title: Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
author:
  - name: Dean Marchiori
    email: deanmarchiori@gmail.com
    attributes:
        corresponding: true
abstract: |
  A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding of associations present in the heterogenous features of the clinical phenotype, and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data on reported cases. This analysis aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To account for censoring in the data, Kaplan-Meier estimation is formulated to calculate patient survival time. The analysis correctly detected the classic phenotype for this condition, as well as a new distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years. This wide range of outcomes is provisionally associated with different variant types however this conclusion could not be validated based on very low sample sizes. An `R` package called `SMPD4` was developed to publish standardised analysis-ready datasets used in this study. This analysis represents the first of its kind to help describe associations and trajectories of individuals with this newly reported condition, despite challenges with sparse and inconsistent data. This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
keywords: 
  - NEDMABA
  - SMPD4  
#journal: "Preprint"    
date: last-modified
execute: 
  echo: false 
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

@@

|                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
|---------------|-------------------------------------|--------------------|
| Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |

: Example of non-tidy free-text descriptions of features {#tbl-nontidy}

| id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
|------------|-------------|------------|------------|------------|------------|
| Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
| Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |

: Example of tidy formatted data where individuals are transposed into rows and text into binary indicators {#tbl-tidy}

@@

To model survival probability for all subjects in the combined data, the data were subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric event indicator which is equal to 1 if the subject is deceased and 0 if the subject was still alive at the time the respective study had concluded. The Kaplan-Meier estimator [@kaplan1958nonparametric] was used to estimate the survival function of the data.

@@

The baseline estimator of all individuals was calculated using the *survival* package in `R` [@survival-book]. Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature. A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* `R` package.

@@

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in red around the origin."
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.2, contrib = NULL),
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

```{r ctr-plot, fig.height=6, fig.width = 8, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  mutate(vars = str_remove(vars, "_[:digit:]")) |> 
  mutate(vars = str_remove(vars, "_ind")) |> 
  mutate(vars = str_replace_all(vars, "_", " ")) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered() +
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  theme(legend.position = 'bottom') +
  labs(y = '')
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

```{r, fig.height=6, fig.width=6,  warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour. A random sample of three labels per cluster are printed to provide a representation of each cluster's characteristics."

@@

  ggplot(d, aes(dim_1, dim_2, colour = clust)) +
  geom_polygon(data = hull, alpha = 0.2, aes(fill = clust)) +
  geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=3), aes(label = var), size = 3, show.legend = F) +
  geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       fill = 'cluster') +
    guides(color='none')
```

@@

```{r}
### Survival
tar_load(survival_data)
tar_load(kapm_baseline)
tar_load(kapm)
```

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98). Survival probability after one year of 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).

```{r}
# KM Curve
x <- survminer::surv_summary(kapm_baseline, data = survival_data)
kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation for the range of recorded survival times, including 95% Confidence Interval.")
```

@@

A comparative box plot of survival times by variant type shows significant outliers for compound heterozygous *missense + splice* and homozygous *missense* variants. However, even within homozygous missense variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.

```{r,fig.width=7, warning=FALSE}
#| label: fig-box
#| fig-cap: "A compariative box plot showing the range of survival times for subjects split by the type of variant. Green bars indicate homozygous variants, while red indicates compound heterozygous variants."

# box plot of variant 
ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
  geom_boxplot() +
  theme(legend.position = 'bottom') +
  labs(fill = 'Variant Type',
       x = 'Survival Time',
       y = '')
```

@@

This meta-review of current research was able to consolidate data that allowed statistical analysis to be conducted on affected individuals. The use of dimension reduction techniques was shown to be successful in uncovering relevant and novel substructure in the data. This was particularly evident with the detection of the accepted typical phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay. After excluding two outlying remarkable cases, we identified the presence of an adjacent cluster of related features that are not well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo. It shows the prolific feature of 'respiratory distress' as more commonly associated with these adjacent features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD). These features are present in the case reported data, but not well described in HPO and help account for the variation seen in the clinical phenotype in the first few principal dimensions.

@@

@@

# References {-}

>>>>

---
title: "Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD"
author: Dean Marchiori
format:
  pdf:
    keep-tex: true
editor: visual
bibliography: bibliography.bib
date: last-modified
execute: 
  echo: false 
---

@@

# Abstract

A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding of associations present in the heterogenous features of the clinical phenotype, and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data on reported cases. This analysis aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To account for censoring in the data, Kaplan-Meier estimation is formulated to calculate patient survival time. The analysis correctly detected the classic phenotype for this condition, as well as a new distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years. This wide range of outcomes is provisionally associated with different variant types however this conclusion could not be validated based on very low sample sizes. An `R` package called `SMPD4` was developed to publish standardised analysis-ready datasets used in this study. This analysis represents the first of its kind to help describe associations and trajectories of individuals with this newly reported condition, despite challenges with sparse and inconsistent data. This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.

@@

```{r}
#| label: tbl-nontidy
#| tbl-cap: Example of non-tidy free-text descriptions of features.
tibble(
  `Family 1- Individual 1` = c("short palpebral fissures, large ears, simple helices"),
  `Family 1- Individual 4 ` = c("short palpebral fissures, receding forehead")
  ) |> 
  gt()
```

```{r}
#| label: tbl-tidy
#| tbl-cap: Example of tidy formatted data where individuals are transposed into rows and text into binary indicators
tribble(
  ~"id", ~"short palpebral fissures", ~"large ears", ~"simple helices", ~"receding forehead", 
  "Family 1- Individual 1", 1, 1, 1, 0,
  "Family 1- Individual 4", 1, 0, 0, 1, 
  ) |> 
  gt()
```

@@

To model survival probability for all subjects in the combined data, the data were subsetted to include `survival_time` which is either the number of days the individual survived for, or the age in days at last follow up, and `deceased` a numeric event indicator which is equal to 1 if the subject is deceased and 0 if the subject was still alive at the time the respective study had concluded. The Kaplan-Meier estimator [@kaplan1958nonparametric] was used to estimate the survival function of the data.

@@

The baseline estimator of all individuals was calculated using the `survival` package in `R` [@survival-book]. Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature. A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the `survival` `R` package.

@@

```{r  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in red around the origin."
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.2, contrib = NULL),
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

```{r ctr-plot, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 5 feature categories as measured by feature contribution from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value."
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  slice_max(n = 5, order_by = contribution, with_ties = F) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  mutate(vars = str_remove(vars, "_[:digit:]")) |> 
  mutate(vars = str_remove(vars, "_ind")) |> 
  mutate(vars = str_replace_all(vars, "_", " ")) |> 
  mutate(vars = str_wrap(vars, width = 25)) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim))) +
  geom_col() +
  scale_y_reordered() +
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  theme(legend.position = 'bottom', axis.text = element_text(size = 8)) +
  labs(y = '', x = '')
```

```{r , warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r ,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

```{r,warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour. A random sample of three labels per cluster are printed to provide a representation of each cluster's characteristics."

@@

  ggplot(d, aes(dim_1, dim_2, colour = clust)) +
  geom_polygon(data = hull, alpha = 0.2, aes(fill = clust)) +
  geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=3), aes(label = var), size = 2, show.legend = F) +
  geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       fill = 'cluster') +
    guides(color='none') +
    theme(legend.position = 'none')
```

@@

```{r}
### Survival
tar_load(survival_data)
tar_load(kapm_baseline)
tar_load(kapm)

x <- survminer::surv_summary(kapm_baseline, data = survival_data)
```

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98). Survival probability after one year of 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52) See @tbl-km.

@@

A comparative box plot of survival times (@fig-box) by variant type shows significant outliers for compound heterozygous *missense + splice* and homozygous *missense* variants. However, even within homozygous missense variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.

```{r, warning=FALSE}
#| label: fig-box
#| fig-cap: "A compariative box plot showing the range of survival times for subjects split by the type of variant. Green bars indicate homozygous variants, while red indicates compound heterozygous variants."

# box plot of variant 
ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
  geom_boxplot() +
  theme(legend.position = 'bottom') +
  scale_fill_grey(start = 0.4, end = 0.9) +
  labs(fill = 'Variant Type',
       x = 'Survival Time',
       y = '')
```

@@

This meta-analysis of current research was able to consolidate data that allowed statistical analysis to be conducted on affected individuals. The use of dimension reduction techniques was shown to be successful in uncovering relevant and novel substructure in the data. This was particularly evident with the detection of the accepted typical phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay. After excluding two outlying remarkable cases, we identified the presence of an adjacent cluster of related features that are not well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo. It shows the prolific feature of 'respiratory distress' as more commonly associated with these adjacent features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD). These features are present in the case reported data, but not well described in HPO and help account for the variation seen in the clinical phenotype in the first few principal dimensions.

@@

```{r}
#| label: tbl-km
#| tbl-cap: "Survival probabilities derived from Kaplan-Meier estimation for the range of recorded survival times, including 95% Confidence Interval."
x |> 
  gt() |> 
  gt::fmt_percent(columns = c(surv, upper, lower)) |> 
  gt::fmt_number(columns = c(std.err), decimals = 2)
```

@@

# References {.unnumbered}
